---
title: "Covariates ([covariates])"
---

The optional `[covariates]` block declares which dataset columns are covariates
and whether each is **continuous** or **categorical**. It is a declaration of
*availability*, not usage — a covariate can be declared without being referenced
in `[individual_parameters]`, and vice versa.

When the block is present, ferx validates the declared columns against the data
and echoes them into the fit's covariate table (`fit$covtab`), enabling
downstream covariate-search and summary workflows.

## Syntax

Two forms are accepted and may be mixed:

```
[covariates]
  WT   continuous
  HT   continuous
  CRCL continuous
  SEX  categorical
  RACE categorical
```

Or the terser `TYPE: NAME, ...` form:

```
[covariates]
  continuous: WT, HT, CRCL
  categorical: SEX, RACE
```

Type keywords are case-insensitive (`continuous`, `cont`, `categorical`, `cat`).
Column names are **case-sensitive** and must match the CSV header exactly.

## Behaviour

| Situation | Effect |
|---|---|
| Block absent | All non-standard CSV columns are auto-detected as covariates (backward-compatible) |
| Column declared, not used in model | Valid — declares type for the covariate table |
| Column used in model, not declared | Model still fits; ferx emits a warning recommending declaration |
| Column declared, absent from data | Hard error: `E_MISSING_COVARIATE` |

## Categorical covariates must be numerically coded

Covariate values are stored as floating-point numbers, so categorical covariates
must be encoded as integer levels in the data (e.g. `SEX` as `0`/`1`, not
`"M"`/`"F"`). When a `[covariates]` block is present this is enforced — a
non-numeric value in a declared covariate is an error (`E_COVARIATE_NOT_NUMERIC`).

Missing values (blank, `.`, `NA`) are permitted and recorded as missing.

## The covariate table (`fit$covtab`)

When `[covariates]` is declared, the fit result carries a covariate table:

```r
fit$covtab          # data frame with ID, TIME, EVID, <declared covariates>
fit$covariate_types # named vector: "continuous" or "categorical" per column
```

Unlike `fit$sdtab` (observation rows only), `fit$covtab` has **one row per
input dataset record**, including dose and other-event rows. Missing values
appear as `NA`. The CLI writes `{model}-covtab.csv` alongside `{model}-sdtab.csv`.

## Example

The bundled `two_cpt_oral_cov` model declares two continuous covariates used
for allometric scaling:

```{r}
#| label: setup
#| message: false
library(ferx)

ex <- ferx_example("two_cpt_oral_cov")
ferx_model_show(ex$model)
```

The relevant block is:

```
[covariates]
  WT   continuous
  CRCL continuous
```

After fitting, inspect the covariate table:

```{r}
#| label: fit
#| cache: true
#| eval: false
fit <- ferx_fit(ex$model, ex$data, verbose = FALSE)

# One row per dataset record (doses + observations)
head(fit$covtab)

# Declared types
fit$covariate_types
```

## Checking dataset columns with `ferx_columns()`

Before writing a model, inspect the CSV header to confirm column names:

```{r}
#| label: columns
ferx_columns(ex)         # pass the ferx_example() list directly
ferx_columns(ex$data)    # or pass the CSV path
```

Use the column names exactly (case-sensitive) in the `[covariates]` block.

## Why declare covariates explicitly?

Without `[covariates]`, every non-standard column is auto-detected, which means:

- Type (continuous vs. categorical) is unknown, so covariate-search tools that
  need to know how to scale a covariate cannot proceed automatically.
- Non-covariate administrative columns (`STUDY`, `DATE`, `SITE`) also appear in
  the covariate table, polluting downstream summaries.
- A typo in `[individual_parameters]` (wrong covariate name) silently evaluates
  to zero rather than raising an error.

Declaring `[covariates]` explicitly gives validation, clean typing, and a focused
covariate table for downstream use.

## See also

- [Individual parameters](individual-parameters.qmd) — using covariates in PK expressions
- [Data selection](data-selection.qmd) — filtering records using covariate columns
- [Example: two-cpt oral with derived AUC](../examples/two-cpt-oral-derived.qmd)
