---
title: "Example: Two-cpt oral with derived AUC"
---

A two-compartment oral PK model with `[derived]` columns for micro-rate
constants and NCA-style summaries: peak (`CMAX`, `TMAX`), first-week AUC
(`AUC_0_168`), and a periodic 24-h AUC (`AUC_TAU`).

## Model

```{r}
#| label: setup
#| message: false
library(ferx)

ex  <- ferx_example("two_cpt_oral_derived")
ferx_model_show(ex$model)
```

The `[derived]` block computes:

| Column | Expression | Description |
|---|---|---|
| `K10` | `CL / V1` | Elimination micro-rate constant |
| `K12` | `Q / V1` | Central→peripheral micro-rate |
| `K21` | `Q / V2` | Peripheral→central micro-rate |
| `CMAX` | `max(IPRED)` | Subject peak concentration |
| `TMAX` | `tmax(IPRED)` | Time of peak |
| `AUC_0_168` | `integral(IPRED, from=0, to=168, step=0.2)` | Week-1 AUC |
| `AUC_TAU` | `integral(IPRED, window=24, anchor=0, step=0.2)` | Periodic 24-h AUC |

`[output]` echoes `CL`, `V1`, `Q`, `V2`, `KA` into sdtab.

::: {.callout-note}
This example uses the `two_cpt_oral_cov` dataset (which has `WT` and `CRCL`
covariate columns). The model does not use these covariates — they are present
in the dataset but ignored by the structural and individual-parameter blocks.
:::

## Fit

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, verbose = FALSE)
print(fit)
```

## Derived columns

```{r}
#| label: columns
names(fit$sdtab)
```

### Micro-rate constants

`K10`, `K12`, `K21` are per-row (constant per subject, dependent on EBE
parameters):

```{r}
#| label: micro-rates
head(unique(fit$sdtab[, c("ID", "CL", "V1", "Q", "V2", "K10", "K12", "K21")]))
```

### Peak and AUC summaries

```{r}
#| label: nca-summary
head(unique(fit$sdtab[, c("ID", "CMAX", "TMAX", "AUC_0_168")]))
```

### Periodic AUC

`AUC_TAU` changes from interval to interval as the concentration profile
accumulates or washes out:

```{r}
#| label: auc-tau
fit$sdtab[fit$sdtab$ID == fit$sdtab$ID[1],
          c("TIME", "TAFD", "AUC_TAU")]
```

## See also

- [Model DSL: Derived columns](../model-dsl/derived.qmd)
- [Example: derived columns](warfarin-derived.qmd)
- [Example: covariate model](covariates.qmd) — same dataset, with covariates
