---
title: "Example: Data selection (LLOQ filtering)"
---

The `[data_selection]` block excludes records from the dataset at read time
without modifying the CSV. It is the ferx equivalent of NONMEM's
`$DATA IGNORE=`. This example drops observations below a surrogate LLOQ of
1.0 mg/L.

## Model

```{r}
#| label: setup
#| message: false
library(ferx)

ex  <- ferx_example("warfarin_data_selection")
ferx_model_show(ex$model)
```

The relevant block is:

```
[data_selection]
  ignore = DV < 1.0
```

## R-side preview

Before fitting, use `ferx_selection()` to inspect which records the filter
would drop:

```{r}
#| label: preview
sel <- ferx_selection(ex$data, ignore = "DV < 1.0")

cat("Total records:", nrow(read.csv(ex$data)), "\n")
cat("Retained:     ", nrow(sel), "\n")
cat("Excluded obs: ", sel$exclusions$n_obs_excluded, "\n")
```

The excluded rows carry a `.exclude_reason` column:

```{r}
#| label: excluded-rows
ferx_selection_excluded(sel)[, c("ID", "TIME", "DV", ".exclude_reason")]
```

## Fit

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, verbose = FALSE)
print(fit)
```

The print output includes a `DATA SELECTION` section showing how many records
were excluded and which conditions fired.

## Exclusion details in `fit$exclusions`

```{r}
#| label: exclusions
fit$exclusions$n_records_total    # total records read
fit$exclusions$n_obs_excluded     # excluded observations
fit$exclusions$fired_ignore       # which ignore conditions fired
```

## Comparison with the unfiltered fit

The two records at `TIME = 120` (the latest time point) are removed. Since
these are the lowest-concentration samples, the effect on estimates is small
but measurable:

```{r}
#| label: compare
#| cache: true
fit_all <- ferx_fit(ferx_example("warfarin")$model,
                    ferx_example("warfarin")$data,
                    verbose = FALSE)

rbind(
  unfiltered = round(c(TVCL = fit_all$theta["TVCL"],
                       TVV  = fit_all$theta["TVV"]), 4),
  filtered   = round(c(TVCL = fit$theta["TVCL"],
                       TVV  = fit$theta["TVV"]), 4)
)
```

## R-side filtering as an alternative to `[data_selection]`

Pass a `ferx_selection()` result directly to `ferx_fit()` instead of writing
a `[data_selection]` block. Both approaches produce the same fit:

```{r}
#| label: r-side
#| eval: false
filtered_data <- ferx_selection(ex$data, ignore = "DV < 1.0")
fit2 <- ferx_fit(ex$model, filtered_data)
```

When both sources supply conditions they are merged — duplicate expressions
are deduplicated automatically.

## See also

- [Model DSL: Data selection](../model-dsl/data-selection.qmd)
- [Fit options](../model-dsl/fit-options.qmd)
