---
title: "Example: ODE with TIME, TAFD, and TAD"
---

Inside `[odes]` RHS expressions four time variables are available:
`TIME` (= `T`) is the ODE solver clock, `TAFD` is time after first dose,
and `TAD` is time after the most recent dose. This example uses them to
build conditional accumulator compartments alongside a standard 1-cpt oral
ODE, and then computes post-fit summaries with `[derived]`.

## Model

```{r}
#| label: setup
#| message: false
library(ferx)

ex  <- ferx_example("warfarin_ode_time")
ferx_model_show(ex$model)
```

Key points:

- `TIME` and `T` are synonymous — both track the ODE solver time axis.
- `TAFD` resets at the first dose; `TAD` resets at each dose event.
- `d/dt(AUC_D1) = if (TIME < 24) central/V else 0.0` accumulates
  AUC only during the first 24 h. The inline `if` fires at every ODE solver
  step, so the integral is exact (no grid approximation needed).
- `d/dt(TAM_INT) = if (central/V > 0.5 && TAD < 24) 1.0 else 0.0` counts
  hours above 0.5 mg/L within each dosing interval. `TAD < 24` resets the
  window at every dose; the condition fires whenever concentration exceeds the
  threshold.
- `[scaling] obs_scale = V` maps the `central` amount to concentration.
- `[derived]` adds `KE`, `T_HALF`, `CMAX`, and `AUC_72` to sdtab.

::: {.callout-note}
**ODE accumulator states vs sdtab.** `AUC_D1` and `TAM_INT` accumulate
correctly during integration and can drive feedback dynamics (e.g., cumulative
exposure affecting tolerance). Raw ODE state values are not written to
`fit$sdtab` automatically. To report an integral in sdtab, use
`[derived] integral(IPRED, ...)` — the two approaches are complementary: use
ODE accumulators when the value feeds back into the RHS, and use `[derived]`
for post-fit sdtab columns.
:::

## Fit

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, verbose = FALSE)
print(fit)
```

## Derived columns

The `[derived]` columns `KE`, `T_HALF`, `CMAX`, and `AUC_72` appear in
sdtab alongside the mandatory minimum:

```{r}
#| label: columns
names(fit$sdtab)
```

```{r}
#| label: derived-vals
head(unique(fit$sdtab[, c("ID", "KE", "T_HALF", "CMAX", "AUC_72")]))
```

## TIME vs TAFD vs TAD in ODE expressions

All three variables are available in `[odes]`. Choosing between them:

| Variable | Resets at | Use for |
|---|---|---|
| `TIME` / `T` | Never | Absolute clock; calendar-time conditions |
| `TAFD` | First dose | Duration since study start |
| `TAD` | Every dose | Within-interval conditions; trough detection |

A common pattern using `TAD` for within-interval integration:

```
d/dt(AUC_interval) = if (TAD < 24) central/V else 0.0
```

And using `TIME` for day-1 vs later conditions:

```
d/dt(AUC_D1) = if (TIME < 24) central/V else 0.0
```

## [scaling] with ODE models

The `obs_scale = V` declaration divides every model prediction by `V` before
the residual error is applied, so the `DV` column should be in
concentration units (amount/volume) while the ODE state `central` is in
amount units. This avoids putting `V` in the ODE RHS for the prediction
step, keeping the ODE equations simpler.

## See also

- [Model DSL: ODE models](../model-dsl/odes.qmd)
- [Model DSL: Derived columns](../model-dsl/derived.qmd)
- [Model DSL: Scaling](../model-dsl/scaling.qmd)
- [Example: ODE model (Michaelis-Menten)](ode-model.qmd)
