---
title: "ADDL/multiple-dose (warfarin)"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Demonstrates the `ADDL` column for multiple-dose designs. A single dose row
per subject (`ADDL=6, II=24`) is expanded to seven daily doses at read time;
`TAD` resets to 0 at each expanded dose automatically. The
[`[derived]`](../model-dsl/derived.qmd) block computes daily trough
(`CMIN_TAU`), peak (`CMAX`), and periodic AUC (`AUC_TAU`).

## Model

```{r model}
library(ferx)
ex <- ferx_example("warfarin_addl")
ferx_model_show(ex$model)
```

## Data structure

The dataset has one dose row per subject with `ADDL=6, II=24` and
observations spread across seven days: rich sampling on day 1, daily troughs
at t = 24, 72, 120, 144 h (where `TAD` resets to 0 at each expanded dose),
and rich steady-state sampling on day 7.

```{r data}
ferx_columns(ex)
head(ex$data[ex$data$EVID == 1, ])   # dose rows: ADDL=6, II=24
```

## Fit

```{r fit}
fit <- ferx_fit(ex$model, ex$data)
print(fit)
```

## Derived columns

```{r sdtab}
# Per-subject summary (scalar derived quantities broadcast to all rows)
subj_summary <- unique(fit$sdtab[, c("ID", "KE", "T_HALF", "CMAX", "CMIN_TAU",
                                      "CMIN_LATE", "AUC_TAU")])
print(subj_summary, row.names = FALSE)
```

`CMIN_TAU` uses `TAD < 1e-10` to locate the observation at the exact start
of each dosing interval (where `TAD` resets to 0 after ADDL expansion).
`CMIN_LATE` uses `TAD > 20` as an alternative window-based trough — less
precise but avoids the floating-point boundary.

```{r day}
# DAY = floor(TAFD / 24) + 1; shows which dosing day each observation belongs to
head(fit$sdtab[, c("ID", "TIME", "TAFD", "TAD", "DAY", "IPRED")], 20)
```

## See also

- [Model DSL: ADDL — multiple additional doses](../model-dsl/steady-state.qmd#addl-multiple-additional-doses)
- [Model DSL: Derived columns](../model-dsl/derived.qmd)
- [Example: Derived columns (warfarin)](warfarin-derived.qmd)
