---
title: "Example: Derived PK/PD quantities"
---

This example uses `[derived]` to compute pharmacodynamic metrics from PK
predictions after the fit: an Emax response, a TAD-gated Cmax, and
time-above-MEC per dosing interval.

The PD parameters (`MEC`, `EMAX`, `EC50`) are **fixed** at assumed values —
they do not enter the PK likelihood and cannot be estimated from PK-only
data. Only `TVCL`, `TVV`, `TVKA`, and the random effects are estimated.

For a model where PD observations are also fitted jointly, see
[Example: simultaneous PK/PD (Emax)](emax-pkpd.qmd).

## Model

```{r}
#| label: setup
#| message: false
library(ferx)

ex  <- ferx_example("warfarin_derived_pkpd")
ferx_model_show(ex$model)
```

The fixed PD thetas appear with `fix` in the `[parameters]` block:

```
theta MEC(0.5)    fix
theta EMAX(80.0)  fix
theta EC50(3.0)   fix
```

The `[derived]` block then uses them to compute:

| Column | Expression | Description |
|---|---|---|
| `EFF` | `EMAX * IPRED / (EC50 + IPRED)` | Emax response at each time point |
| `CMAX_INTERVAL` | `max(IPRED, TAD < 24)` | Peak in current dosing interval |
| `TAM_TAU` | `integral(1.0, IPRED > MEC, window=24, anchor=0, step=0.1)` | Hours above MEC per 24-h interval |
| `TAM_D1` | `integral(1.0, IPRED > MEC, from=0, to=24, step=0.1)` | Hours above MEC on day 1 |

## Fit

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, verbose = FALSE)
print(fit)
```

The PK thetas have proper SEs; `MEC`, `EMAX`, and `EC50` show `SE = 0`
because they are fixed:

```{r}
#| label: estimates
ferx_estimates(fit)
```

## PD derived columns

```{r}
#| label: pd-cols
head(fit$sdtab[, c("ID", "TIME", "IPRED", "EFF", "TAM_TAU", "TAM_D1",
                   "CMAX_INTERVAL")], 12)
```

`EFF` varies per time point (it depends on `IPRED`). `CMAX_INTERVAL`,
`TAM_TAU`, and `TAM_D1` are aggregate quantities — constant per 24-h window
or per subject.

## Why `step=` for time-above-threshold

`TAM_TAU` uses `step=0.1` to force grid evaluation at 0.1-hour intervals.
Without `step=`, only observation time points are evaluated. If the
observation design is sparse (large gaps between samples), the trapezoidal
rule can miss brief periods above threshold between those times. The `step=`
argument fills in the gaps:

```
# Without step= (observation-time only -- can miss threshold crossings):
TAM_approx = integral(1.0, IPRED > MEC, window=24, anchor=0)

# With step=0.1 (fine grid -- catches brief excursions):
TAM_TAU    = integral(1.0, IPRED > MEC, window=24, anchor=0, step=0.1)
```

## See also

- [Model DSL: Derived columns](../model-dsl/derived.qmd)
- [Example: derived columns](warfarin-derived.qmd) — full [derived] syntax tour
- [Example: simultaneous PK/PD (Emax)](emax-pkpd.qmd) — jointly estimated PD
