---
title: "Example: Derived columns ([derived] and [output])"
---

The `[derived]` block adds computed columns to `sdtab` after the fit.
The `[output]` block echoes individual PK parameters and covariates into
`sdtab` that would not be there otherwise. Neither block affects estimation —
they are pure post-processing.

## Model

```{r}
#| label: setup
#| message: false
library(ferx)

ex  <- ferx_example("warfarin_derived")
ferx_model_show(ex$model)
```

The model adds eleven derived columns covering all three computation kinds:

| Column | Kind | Description |
|---|---|---|
| `KE`, `T_HALF`, `DAY`, `TAU_TIME` | Per-row | Evaluated at each observation time |
| `CMAX`, `TMAX`, `CTROUGH`, `CMAX_D1`, `CMAX_D14` | Aggregate | One value per subject, broadcast to all rows |
| `AUC_0_72`, `AUC_TAU`, `AUC_DV_72` | Integral | Trapezoidal / periodic / DV-based AUC |

`[output]` then adds `CL`, `V`, and `KA` (individual EBE parameter values).

## Fit

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, verbose = FALSE)
print(fit)
```

## sdtab columns

```{r}
#| label: columns
names(fit$sdtab)
```

## Per-row columns

`KE` and `T_HALF` depend only on the subject's EBE `CL`/`V`, so they are
constant per subject but appear on every row. `DAY` increments with each
24-h window.

```{r}
#| label: perrow
head(fit$sdtab[, c("ID", "TIME", "TAFD", "KE", "T_HALF", "DAY")], 12)
```

## Aggregate columns

`CMAX`, `TMAX`, and `AUC_0_72` are the same on every row of a subject:

```{r}
#| label: aggregate
unique(fit$sdtab[, c("ID", "CMAX", "TMAX", "AUC_0_72")])
```

## Trough with `TAD < 1e-10`

`CTROUGH = min(IPRED, TAD < 1e-10)` selects the individual prediction at
dose time. The `1e-10` guard is necessary because floating-point modular
arithmetic can produce `TAD = 1e-13` rather than exactly `0`, so `TAD == 0`
is unreliable:

```{r}
#| label: ctrough
unique(fit$sdtab[, c("ID", "CTROUGH")])
```

## Periodic AUC (`AUC_TAU`)

`AUC_TAU = integral(IPRED, window=24, anchor=0, step=0.1)` produces one
value per 24-h dosing window. It changes across windows as the concentration
profile evolves:

```{r}
#| label: auc-tau
fit$sdtab[fit$sdtab$ID == fit$sdtab$ID[1],
          c("TIME", "TAFD", "AUC_TAU")]
```

## DV-based AUC (`AUC_DV_72`)

`AUC_DV_72 = integral(DV, from=0, to=72)` uses the **observed values** at
each observation time rather than IPRED. The `step=` argument is ignored for
DV-based integrals (no interpolation between observations):

```{r}
#| label: auc-dv
unique(fit$sdtab[, c("ID", "AUC_0_72", "AUC_DV_72")])
```

## [output] — individual parameter columns

`CL`, `V`, and `KA` appear in `sdtab` because of the `[output]` block.
Without it, only the mandatory minimum columns plus `[derived]` names would
be present:

```{r}
#| label: output-cols
unique(fit$sdtab[, c("ID", "CL", "V", "KA")])
```

## See also

- [Model DSL: Derived columns](../model-dsl/derived.qmd)
- [Example: derived PK/PD quantities](warfarin-derived-pkpd.qmd)
- [Example: two-cpt oral with derived AUC](two-cpt-oral-derived.qmd)
